Supermartingales and Termination Analysis of Statistical PCF

### 2021-01-18 LICS21/supermartingales.tex
Minor fixes

It will be demonstrated later that for any rankable term $M$, if $(M_n)_{n \geq 0}$ \changed[lo]{(qua stochastic process)} is the reduction sequence starting from $M$, then $(f(M_n))_{n \geq 0}$ is a supermartingale, and $M$ terminates almost surely, but first, some preliminaries about supermartingales (see e.g.~\cite{Williams1999} for details).
\iffalse
\lo{It is confusing (strictly speaking, incorrect) to say that $(M_n)_{n \geq 0}$ is a reduction sequence from $M$. Formally $M_n$ is the random variable $M_n: s \mapsto \pi_0 (\red^n(M, s))$.}
\akr{I guess it's implicitly identifying a sequence of functions ($(M_n)_{n \geq 0}$), and a function producing sequences (a reduction-sequence-valued random variable), but this sort of thing seems very standard in dealing with random variables (e.g.~applying a function to some random variables, which is really a sort of concatenation).}
\fi
%\paragraph{}

@@

\begin{definition}\rm
\begin{asparaenum}[(i)]
\item A sequence of r.v.~$(Y_n)_{n \geq 0}$ adapted to a filtration $(\calF_n)_{n \geq 0}$ is a \emph{supermartingale} if for all $n \geq 0$, $Y_n$ is integrable (i.e.~$\expect{|Y_n|} < \infty$), and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n$ a.s.~(i.e.~for all $A \in \calF_n$, $\int_A \mathbb{P}(\dif \omega) \, Y_{n+1}(\omega) \leq \int_A \mathbb{P}(\dif \omega) \,Y_n(\omega)$).
\item Let $\epsilon > 0$. 
Given a stopping time $T$ and a supermartingale $(Y_n)_{n \geq 0}$, both adapted to filtration $(\calF_n)_{n \geq 0}$, 
we say that $(Y_n)_{n \geq 0}$ is a \emph{$\epsilon$-ranking supermartingale w.r.t.~$T$} if for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$.
\footnote{For $A \in \calF$, we write ${\bf 1}_A$ for the \emph{indicator function} of $A$, i.e., the random variable defined by ${\bf 1}_A(\omega) := 1$ if $\omega \in A$, and $0$ otherwise.} %\lo{\LaTeX\ does not allow the preceding bracketed sentence to be typeset as a footnote. Strange!}\lo{Now it does. Stranger.}
A \emph{ranking supermartingale} w.r.t.~$T$ is a $\epsilon$-ranking supermartingale w.r.t.~$T$ for some $\epsilon > 0$.
\end{asparaenum}
\end{definition}

@@

We first state a useful property about r.v.~$T_0, T_1, T_2, \cdots$.
\begin{restatable}{lemma}{TnBounded}
\label{lem:TnBounded}
$(T_n)_{n \geq 0}$ is an increasing sequence of stopping times adapted to $(\calF_n)_{n \geq 0}$, and each $T_i$ is bounded.
\end{restatable}

>>>>

It will be demonstrated later that for any rankable term $M$, if $(M_n)_{n \geq 0}$ is the reduction sequence starting from $M$ (considered as a stochastic process), then $(f(M_n))_{n \geq 0}$ is a supermartingale, and $M$ terminates almost surely, but first, some preliminaries about supermartingales (see e.g.~\cite{Williams1999} for details).
\iffalse
\lo{It is confusing (strictly speaking, incorrect) to say that $(M_n)_{n \geq 0}$ is a reduction sequence from $M$. Formally $M_n$ is the random variable $M_n: s \mapsto \pi_0 (\red^n(M, s))$.}
\akr{I guess it's implicitly identifying a sequence of functions ($(M_n)_{n \geq 0}$), and a function producing sequences (a reduction-sequence-valued random variable), but this sort of thing seems very standard in dealing with random variables (e.g.~applying a function to some random variables, which is really a sort of concatenation).}
\fi
%\paragraph{}

@@

\begin{definition}\rm
\begin{asparaenum}[(i)]
\item A sequence of r.v.s $(Y_n)_{n \geq 0}$ adapted to a filtration $(\calF_n)_{n \geq 0}$ is a \emph{supermartingale} if for all $n \geq 0$, $Y_n$ is integrable (i.e.~$\expect{|Y_n|} < \infty$), and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n$ a.s.~(i.e.~for all $A \in \calF_n$, $\int_A \mathbb{P}(\dif \omega) \, Y_{n+1}(\omega) \leq \int_A \mathbb{P}(\dif \omega) \,Y_n(\omega)$).
\item Let $\epsilon > 0$. 
Given a stopping time $T$ and a supermartingale $(Y_n)_{n \geq 0}$, both adapted to filtration $(\calF_n)_{n \geq 0}$, 
we say that $(Y_n)_{n \geq 0}$ is a \emph{$\epsilon$-ranking supermartingale w.r.t.~$T$} if for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$.
\footnote{For $A \in \calF$, we write ${\bf 1}_A$ for the \emph{indicator function} of $A$, i.e., the random variable defined by ${\bf 1}_A(\omega) := 1$ if $\omega \in A$, and $0$ otherwise.} %\lo{\LaTeX\ does not allow the preceding bracketed sentence to be typeset as a footnote. Strange!}\lo{Now it does. Stranger.}
A \emph{ranking supermartingale} w.r.t.~$T$ is a $\epsilon$-ranking supermartingale w.r.t.~$T$ for some $\epsilon > 0$.
\end{asparaenum}
\end{definition}

@@

We first state a useful property about r.v.s $T_0, T_1, T_2, \dots$.
\begin{restatable}{lemma}{TnBounded}
\label{lem:TnBounded}
$(T_n)_{n \geq 0}$ is an increasing sequence of stopping times adapted to $(\calF_n)_{n \geq 0}$, and each $T_i$ is bounded.
\end{restatable}
